---
title: "Born Again and Bayesian tree approaches"
author: "Marjolein Fokkema"
date: "17-7-2022"
output: pdf_document
---

Many have found artificial data-generation effective to improve the performance of more simple / unstable models.

Instability generally reduced with larger sample size and higher signal-to-noise ratio. 

By using accurate black-box, can generate data with 



How to generate $X$?

1: Keep X exactly as is ($N_{gen} = 1$. 
2: Randomly permute subset of values (smearing)
3: Simply repeat observed data

How to generate $Y$?

1: Deterministic, given $X_{gen}$.
2: From Bayesian posterior distribution.


What works: 

Born-again approach for treatment effects, without permuting $X$.

Born-again approach for safety data.

What does not work:

Born-again approach for longitudinal datasets; ECLSK, ACT.


```{r}
## Bayesian data generation function
## Returns X_gen, y_gen
gendata_bay <- function(data, PPD, ngen) {
  
  if (nrow(data) != ncol(PPD)) 
    stop("Amount of observations in PPD is different from data")
  
  index_gen <- sample.int(n = nrow(data), size = ngen*nrow(data), replace = TRUE)
  x_gen <- data[index_gen, ]
  
  ## Posterior Predictive Distribution has 2000 rows (default) and nobs columns
  ## Thus, for each bootstrap sampled row, selected correspdonding column
  ## and sample a random value from that posterior
  y_gen <- sapply(index_gen, function(i) sample(x = PPD[ , i], size = 1))
  
  return(cbind(x_gen, y_gen))
}


## Smearing data generation function
## Returns X_gen only
gendata_smr <- function(data, palt = 0.5, ngen) {

  if (palt < 0 | palt > 1) stop("palt must be bound between between 0 and 1")
  
  index_gen <- sample.int(n = nrow(data), size = ngen*nrow(data), replace = TRUE)
  gen_data <- data[index_gen, ]
  
  ## With probability palt, replace value in X_gen with random sample from 
  ## corresponding column
  if (palt != 0) {
    for (i in 1:nrow(gen_data)) {
      for (j in 1:ncol(gen_data)) {
        if (rbinom(n = 1, size = 1, prob = palt)) { 
          gen_data[i, j] <- sample(x = data[ , j], size = 1)
        }
      }
    }
  }
  return(gen_data)
}
```


# Replicate Breiman analyses

Datasets analysed by Breiman:

Classification: breast cancer, ionosphere, glass, soybean, sonar

Regression: Boston Housing, Ozone, Friedman #1, #2, #3 (200 train, 2000 test)

## Boston Housing

```{r, eval=FALSE}
library("glmertree")
library("dbarts")
library("mlbench")

data(BostonHousing)
data <- BostonHousing

nfolds <- 10
nreps <- 10
ngensizes <- c(1, 5, 10)
palts <- c(0, 0.25, 0.5, 1.0)

## Create objects for saving results (MSE for Bayesian approach, MSE2 for Breiman approach)
MSE <- data.frame(matrix(data = NA, nrow = nfolds*nreps, 
                         ncol = 3 + length(ngensizes) + length(ngensizes)*length(palts)))
BaSmear_design <- expand.grid(ngensizes, palts)
colnames(BaSmear_design) <- c("ngen", "palt")
colnames(MSE) <- c("GLMM_tree", "Bart", "Ba", 
                   paste0("BaBayes_N=", ngensizes),
                   paste0("BaSmear_N=", BaSmear_design[ , 1], "_palt=", BaSmear_design[ , 2]))
rownames(MSE) <- paste0("rep", sort(rep(1:nreps, times = nfolds)), "_fold", rep(1:nfolds, times = nreps))
#MSE <- MSE[ , -which(is.na(colnames(MSE)))]
tree_size <- MSE

set.seed(42)
for (k in 1:nreps) {
    
  ## Generate train-test splits
  fold_ids <- sample(rep(1:nfolds, times = ceiling(nrow(data)/nfolds)), 
                     size = nrow(data), replace = FALSE)
  
  for (i in 1:nfolds) {
    
    ## Select train and test data
    traindat <- data[fold_ids != i, ]
    testdat <- data[fold_ids == i, ]
    
    ## Fit regular GLMM tree
    gt <- lmtree(medv ~ 1 | ., data = traindat)
    pred_gt <- predict(gt, newdata = testdat)
    MSE[(k-1)*10+i, 1] <- mean((testdat$medv - pred_gt)^2)
    tree_size[(k-1)*10+i, 1] <- (length(gt)-1)/2

    
    ## 1) Fit (and evaluate) BART
    ## 2) Generate surrogate data
    ## 3) Fit and evaluate tree fitted to surrogate data
    
    ## Fit Bayesian born-again approach
    ##
    brt <- bart2(medv ~ ., data = traindat, test = testdat, n.trees = 100, 
                 combineChains = TRUE, verbose = FALSE)
    MSE[(k-1)*10+i, 2] <- mean((testdat$medv - brt$yhat.test.mean)^2)
    surr_dat <- traindat
    surr_dat$medv <- brt$yhat.train.mean
    ba_gt <- lmtree(medv ~ 1 | ., data = surr_dat)
    pred_babart <- predict(ba_gt, newdata = testdat)
    MSE[(k-1)*10+i, 3] <- mean((testdat$medv - pred_babart)^2)
    tree_size[(k-1)*10+i, 3] <- (length(ba_gt)-1)/2
    
    posterior <- brt$yhat.train
    for (ngen in ngensizes) {
      surr_dat <- gendata_bay(data = traindat, PPD = posterior, ngen = ngen)
      ba_gt <- lmtree(y_gen ~ 1 | . - medv, data = surr_dat,
                      weights = rep(1/ngen, times = nrow(surr_dat)))
      pred_babart <- predict(ba_gt, newdata = testdat)
      MSE[(k-1)*10+i, 3+match(ngen, ngensizes)] <- mean((testdat$medv - pred_babart)^2)
      tree_size[(k-1)*10+i, 3+match(ngen, ngensizes)] <- (length(ba_gt)-1)/2
    }
       
    ## Fit Breiman born-again approach
    ##
    for (ngen_palt in 1:nrow(BaSmear_design)) {
      palt <- BaSmear_design$palt[ngen_palt]
      ngen <- BaSmear_design$ngen[ngen_palt]
      surr_dat <- gendata_smr(data = traindat, palt = palt, ngen = ngen)
      brt <- bart2(medv ~ ., data = traindat, test = surr_dat, verbose = FALSE,
                   n.trees = 100, combineChains = TRUE)
      y_gen <- brt$yhat.test.mean
      surr_dat <- cbind(surr_dat, y_gen)
      ba_gt <- lmtree(y_gen ~ 1 | . -medv, data = surr_dat,
                       weights = rep(1/ngen, times = nrow(surr_dat)))
      pred_babart <- predict(ba_gt, newdata = testdat)
      MSE[(k-1)*10+i, match(ngen, ngensizes) + match(palt, palts)*length(ngensizes) + 3] <- 
        mean((testdat$medv - pred_babart)^2)
      tree_size[(k-1)*10+i, match(ngen, ngensizes) + match(palt, palts)*length(ngensizes) + 3] <- 
        (length(ba_gt)-1)/2
    }
    
    print(paste0("Finished fold ", i, " of repetition ", k))
  }
}
save(MSE, file = "BostonHousing MSE.Rda")
save(tree_size, file = "BostonHousing tree_size.Rda")
``` 

```{r}
load(file = "BostonHousing MSE.Rda")
load(file = "BostonHousing tree_size.Rda")
boxplot(MSE)
colMeans(MSE)
which.min(colMeans(MSE[ , -2]))
sapply(MSE, sd)
boxplot(tree_size)
colMeans(tree_size)
sapply(tree_size, sd)
```


## Ozone

```{r}
library("glmertree")
library("dbarts")
library("mlbench")
library("partykit")

data(Ozone)
sapply(Ozone, function(x) table(is.na(x)))
data <- Ozone[ , -9]
data <- na.omit(data)

nfolds <- 10
nreps <- 10
ngensizes <- c(1, 5, 10)
palts <- c(0, 0.25, 0.5, 1.0)

## Create objects for saving results (MSE for Bayesian approach, MSE2 for Breiman approach)
MSE <- data.frame(matrix(data = NA, nrow = nfolds*nreps, 
                         ncol = 3 + length(ngensizes) + length(ngensizes)*length(palts)))
BaSmear_design <- expand.grid(ngensizes, palts)
colnames(BaSmear_design) <- c("ngen", "palt")
colnames(MSE) <- c("GLMM_tree", "Bart", "Ba", 
                   paste0("BaBayes_N=", ngensizes),
                   paste0("BaSmear_N=", BaSmear_design[ , 1], "_palt=", BaSmear_design[ , 2]))
rownames(MSE) <- paste0("rep", sort(rep(1:nreps, times = nfolds)), "_fold", rep(1:nfolds, times = nreps))
#MSE <- MSE[ , -which(is.na(colnames(MSE)))]
tree_size <- MSE

set.seed(42)
for (k in 1:nreps) {
    
  ## Generate train-test splits
  fold_ids <- sample(rep(1:nfolds, times = ceiling(nrow(data)/nfolds)), 
                     size = nrow(data), replace = FALSE)
  
  for (i in 1:nfolds) {
    
    ## Select train and test data
    traindat <- data[fold_ids != i, ]
    testdat <- data[fold_ids == i, ]
    
    ## Fit regular GLMM tree
    gt <- lmtree(V4 ~ 1 | ., data = traindat)
    pred_gt <- predict(gt, newdata = testdat)
    MSE[(k-1)*10+i, 1] <- mean((testdat$V4 - pred_gt)^2)
    tree_size[(k-1)*10+i, 1] <- (length(gt)-1)/2
    
    ## 1) Fit (and evaluate) BART
    ## 2) Generate surrogate data
    ## 3) Fit and evaluate tree fitted to surrogate data
    
    ## Fit Bayesian born-again approach
    brt <- bart2(V4 ~ ., data = traindat, test = testdat, n.trees = 100, 
                 combineChains = TRUE, verbose = FALSE)
    MSE[(k-1)*10+i, 2] <- mean((testdat$V4 - brt$yhat.test.mean)^2)
    surr_dat <- traindat
    surr_dat$V4 <- brt$yhat.train.mean
    ba_gt <- lmtree(V4 ~ 1 | ., data = surr_dat)
    pred_babart <- predict(ba_gt, newdata = testdat)
    MSE[(k-1)*10+i, 3] <- mean((testdat$V4 - pred_babart)^2)
    tree_size[(k-1)*10+i, 3] <- (length(ba_gt)-1)/2
    posterior <- brt$yhat.train
    for (ngen in ngensizes) {
      surr_dat <- gendata_bay(data = traindat, PPD = posterior, ngen = ngen)
      ba_gt <- lmtree(y_gen ~ 1 | . - V4, data = surr_dat,
                      weights = rep(1/ngen, times = nrow(surr_dat)))
      pred_babart <- predict(ba_gt, newdata = testdat)
      MSE[(k-1)*10+i, 3+match(ngen, ngensizes)] <- mean((testdat$V4 - pred_babart)^2)
      tree_size[(k-1)*10+i, 3+match(ngen, ngensizes)] <- (length(ba_gt)-1)/2
    }
       
    ## Fit Breiman born-again approach
    for (ngen_palt in 1:nrow(BaSmear_design)) {
      palt <- BaSmear_design$palt[ngen_palt]
      ngen <- BaSmear_design$ngen[ngen_palt]
      surr_dat <- gendata_smr(data = traindat, palt = palt, ngen = ngen)
      brt <- bart2(V4 ~ ., data = traindat, test = surr_dat, verbose = FALSE,
                   n.trees = 100, combineChains = TRUE)
      y_gen <- brt$yhat.test.mean
      surr_dat <- cbind(surr_dat, y_gen)
      ba_gt <- lmtree(y_gen ~ 1 | . -V4, data = surr_dat,
                       weights = rep(1/ngen, times = nrow(surr_dat)))
      pred_babart <- predict(ba_gt, newdata = testdat)
      MSE[(k-1)*10+i, match(ngen, ngensizes) + match(palt, palts)*length(ngensizes) + 3] <- 
        mean((testdat$V4 - pred_babart)^2)
      tree_size[(k-1)*10+i, match(ngen, ngensizes) + match(palt, palts)*length(ngensizes) + 3] <- 
        (length(ba_gt)-1)/2
    }
    
    print(paste0("Finished fold ", i, " of repetition ", k))
  }
}
save(MSE, file = "Ozone MSE.Rda")
save(tree_size, file = "Ozone tree_size.Rda")
``` 

```{r}
load(file = "Ozone MSE.Rda")
load(file = "Ozone tree_size.Rda")
boxplot(MSE)
colMeans(MSE)
which.min(colMeans(MSE[ , -2]))
sapply(MSE, sd)
boxplot(tree_size)
colMeans(tree_size)
sapply(tree_size, sd)
```

## Friedman

```{r}
library("dbarts")
library("mlbench")

data(BostonHousing)
data <- BostonHousing
#Ozone
#mlbench.friedman1(200)
```

# Clustered data that is not repeated measures:

https://uclspp.github.io/POLS0010/7-multilevel-regression-and-post-stratification-mrp.html
